---
title: "Practical 5 Answers"
subtitle: "GLMs: Model comparisons"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    link-external-newwindow: true
    link-external-icon: true
editor: source
---

## Getting started

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(readr)
library(easystats) # for geom_violinhalf function

melanoma <- read_csv("data/melanoma.csv")
```

## The melanoma dataset

### Question 1

Load the data and prepare it for logistic regression. Recode status, sex and ulcer as properly labelled factors. Create a new variable called status_binary from status to record if the patient died from melanoma or not (this variable should be binary).

```{r}
#| label: data-cleaning

melanoma <- read_csv("data/melanoma.csv") |>
  mutate(
    status = factor(status, levels = 1:3,
                    labels = c("died - melanoma", "alive", "died - other")),
    sex = factor(sex, levels = 0:1, labels = c("female", "male")),
    ulcer = factor(ulcer, levels = 0:1, labels = c("not present", "present")),
    status_binary = if_else(status == "died - melanoma", 1, 0)
  )

melanoma
```

> `status_binary` recodes the three-level `status` into a binary outcome.
>
> -   `1` if the patient died from melanoma
> -   `0` otherwise, meaning either alive or died from another cause.
>
> This simplification/limitation should always be noted when discussing the analysis approach.

### Question 2

Produce exploratory plots examining the relationship between status_binary and each of the candidate predictors. Select appropriate plots for each variable type. Consider violin plots, bar charts, etc.

```{r}
#| label: explore-thickness

melanoma |>
  ggplot(aes(x = factor(status_binary, labels = c("no", "yes")), y = thickness)) +
  geom_violinhalf(fill = "grey85") +
  geom_jitter(width = 0.1, alpha = 0.4) +
  scale_x_discrete("Died from melanoma?") +
  scale_y_continuous("Tumour thickness (mm)") +
  theme_bw()
```

```{r}
#| label: explore-age

melanoma |>
  ggplot(aes(x = factor(status_binary, labels = c("no", "yes")), y = age)) +
  geom_violinhalf(fill = "grey85") +
  geom_jitter(width = 0.1, alpha = 0.4) +
  scale_x_discrete("Died from melanoma?") +
  scale_y_continuous("Age at operation (years)") +
  theme_bw()
```

```{r}
#| label: explore-sex

melanoma |>
  ggplot(aes(x = sex, fill = factor(status_binary, labels = c("no", "yes")))) +
  geom_bar(position = "fill") +
  scale_x_discrete("Sex") +
  scale_y_continuous("Proportion") +
  scale_fill_discrete("Died from melanoma?") +
  theme_bw()
```

```{r}
#| label: explore-ulcer

melanoma |>
  ggplot(aes(x = ulcer, fill = factor(status_binary, labels = c("no", "yes")))) +
  geom_bar(position = "fill") +
  scale_x_discrete("Ulceration") +
  scale_y_continuous("Proportion") +
  scale_fill_discrete("Died from melanoma?") +
  theme_bw()
```

```{r}
#| label: explore-year

melanoma |>
  ggplot(aes(x = factor(status_binary, labels = c("no", "yes")), y = year)) +
  geom_boxplot() +
  scale_x_discrete("Died from melanoma?") +
  scale_y_continuous("Year of operation") +
  theme_bw()
```

### Question 3

Based on your plots, which predictors do you expect to be associated with status_binary? Do the relationships make clinical sense?

> The plots suggest `thickness` and `ulcer` are the two predictors most clearly associated with `status_binary`. Patients who died from melanoma have visibly thicker tumours on average, and a noticeably larger share of the “died from melanoma” group has `ulcer = present`.
>
> There could also be an effect from biological `sex`, which shows a mild association. In contrast, `age` and `year` show little visible separation between the two outcome groups.
>
> This makes good clinical sense: tumour thickness (Breslow depth) and ulceration are two established prognostic markers for melanoma.

## Fitting binary GLMs

### Question 4

Fit the following logistic regression model, where (y_i) is the `status_binary` variable:

$$
M_3:\quad \operatorname{logit}(y_i)
= \beta_0 + \beta_1\text{thickness}_i
+ \beta_2\text{sex}_i
+ \beta_3\text{age}_i
$$

```{r}
#| label: binary-glm

M3 <- glm(status_binary ~ thickness + sex + age,
          data = melanoma,
          family = binomial())

summary(M3)
```

### Question 5

What are the estimated log-odds for thickness? Interpret this on the odds-ratio scale (exponentiate the coefficient and provide a confidence interval using confint()).

```{r}
#| label: thickness-coefficient

coef(M3)["thickness"]
```

```{r}
#| label: thickness-odds-ratio

exp(coef(M3)["thickness"])
```

```{r}
#| label: thickness-confidence-interval

exp(confint(M3)["thickness", ])
```

> Interpretations:
>
> -   **Log-odds scale:** the log-odds coefficient for `thickness` is `0.187`. Each additional 1 mm of tumour thickness increases the log-odds of dying from melanoma by `0.187`, holding `sex` and `age` fixed.
>
> -   **Odds-ratio scale:** the odds ratio for `thickness` is `1.206`. Each additional 1 mm of tumour thickness is associated with about a **21% increase** in the odds of dying from melanoma, holding `sex` and `age` fixed.
>
> -   The 95% confidence interval is approximately `1.08` to `1.36`. Since the interval does not include `1`, there is a statistically significant positive association at the 5% level.

### Question 6

Is age a statistically significant predictor in M3 at the 5% level, adjusting for thickness and sex? What test is being used here?

```{r}
#| label: age-test

summary(M3)$coefficients["age", ]
```

> `age` has a test statistic of approximately `0.512` with a p-value of `0.609`.
>
> Since `p = 0.609 > 0.05`, the association between `age` and `status_binary` is not statistically significant at the 5% level.
>
> This is a **Wald test**, which compares the estimated coefficient with its standard error to assess the strength of association between the outcome and the covariate.

## Likelihood ratio tests

### Question 7

Fit these further regression models.

$$
M_0: \operatorname{logit}(y_i) = \beta_0
$$

$$
M_1: \operatorname{logit}(y_i) = \beta_0 + \beta_1 \text{thickness}_i
$$

$$
M_2: \operatorname{logit}(y_i) = \beta_0 + \beta_1 \text{thickness}_i + \beta_2 \text{sex}_i
$$

```{r}
#| label: nested-models

M0 <- glm(status_binary ~ 1,
          data = melanoma,
          family = binomial())

M1 <- glm(status_binary ~ thickness,
          data = melanoma,
          family = binomial())

M2 <- glm(status_binary ~ thickness + sex,
          data = melanoma,
          family = binomial())
```

### Question 8

The models $M_0 \subset M_1 \subset M_2 \subset M_3$ form a nested sequence. Use `anova()` with `test = "LRT"` to test each step of this sequence.

i. For each step, state the null and alternative hypotheses.

ii. Which predictors contribute a statistically significant improvement in fit at $\alpha = 0.05$?

iii. The LRT statistic for comparing $M_k$ to $M_{k-1}$ is

$$
D = 2(\ell_{M_k} - \ell_{M_{k-1}}),
$$

where $\ell$ denotes the log-likelihood. Verify the test statistic for the $M_0$ vs $M_1$ comparison by extracting the log-likelihoods manually using `logLik()`.

```{r}
#| label: likelihood-ratio-test

anova(M0, M1, M2, M3, test = "LRT")
```

> **(i) Null and alternative hypotheses for each step:**
>
> -   `M0` vs `M1`: Does adding `thickness` improve the intercept-only model?
>
> -   `M1` vs `M2`: Does adding `sex` improve the model, given that `thickness` is already included?
>
> -   `M2` vs `M3`: Does adding `age` improve the model, given that `thickness` and `sex` are already included?
>
> **(ii) Which steps are significant at α = 0.05?**
>
> Only the first comparison is statistically significant.
>
> Adding `thickness` to the null model produces a statistically significant improvement in fit (`χ² = 16.2227`, `df = 1`, `p = 5.632 × 10⁻⁵`).
>
> Adding `sex` is not significant (`p = 0.1204`), and adding `age` is also not significant (`p = 0.6076`).
>
> These results are consistent with the previous Wald test and the EDA plots.
>
> **(iii) Manual verification of the `M0` vs `M1` statistic:**

```{r}
#| label: manual-lrt

logLik(M0)
```

```{r}
logLik(M1)
```

```{r}
D <- 2 * (as.numeric(logLik(M1)) - as.numeric(logLik(M0)))
D
```

```{r}
pchisq(D, df = 1, lower.tail = FALSE)
```

> This gives `D = 16.22268` and `p = 5.631574 × 10⁻⁵`, matching the `anova()` output.

## AIC

### Question 9

Perform backward stepwise model selection with AIC manually, beginning with the following model. What is the final model? Does this differ from the previous choice?

$$
M_4:
\operatorname{logit}(y_i)
=
\beta_0
+ \beta_1\,\texttt{thickness}_i
+ \beta_2\,\texttt{sex}_i
+ \beta_3\,\texttt{age}_i
+ \beta_4\,\texttt{ulcer}_i
$$

```{r}
#| label: aic-model

M4 <- glm(status_binary ~ thickness + sex + age + ulcer,
          data = melanoma,
          family = binomial())
```

> We can create new models explicitly dropping the term that is not required. But there is a shortcut in R using `update`.
>
> The `M4` model with `thickness` removed can be created with:
>
> `M4_without_thickness <- update(M4, . ~ . - thickness)`
>
> If we choose not to save these intermediate models then the following code works.

```{r}
#| label: aic-step1

aic_step1 <- list()

# Reference AIC
aic_step1$initial <- AIC(M4)

# Starting from M4, try dropping each term
aic_step1$thickness <- AIC(update(M4, . ~ . - thickness))
aic_step1$sex <- AIC(update(M4, . ~ . - sex))
aic_step1$age <- AIC(update(M4, . ~ . - age))
aic_step1$ulcer <- AIC(update(M4, . ~ . - ulcer))

aic_step1
```

> The AIC values are:
>
> -   Full model: `220.31`
> -   Without `thickness`: `222.02`
> -   Without `sex`: `219.68`
> -   Without `age`: `218.46`
> -   Without `ulcer`: `231.45`
>
> Dropping `age` gives the lowest AIC (`218.46`), so we drop `age` and continue from `status_binary ~ thickness + sex + ulcer`.

```{r}
#| label: aic-step2

model_after_step1 <- glm(
  status_binary ~ thickness + sex + ulcer,
  data = melanoma,
  family = binomial()
)

aic_step2 <- list()

aic_step2$initial <- AIC(model_after_step1)
aic_step2$thickness <- AIC(update(model_after_step1, . ~ . - thickness))
aic_step2$sex <- AIC(update(model_after_step1, . ~ . - sex))
aic_step2$ulcer <- AIC(update(model_after_step1, . ~ . - ulcer))

aic_step2
```

> Dropping `sex` yields the maximum improvement for AIC (`217.85`, vs. `218.46`). Continue from `status_binary ~ thickness + ulcer`.

```{r}
#| label: aic-step3

model_after_step2 <- glm(
  status_binary ~ thickness + ulcer,
  data = melanoma,
  family = binomial()
)

aic_step3 <- list()

aic_step3$initial <- AIC(model_after_step2)
aic_step3$thickness <- AIC(update(model_after_step2, . ~ . - thickness))
aic_step3$ulcer <- AIC(update(model_after_step2, . ~ . - ulcer))

aic_step3
```

> Both remaining removals increase AIC, so we stop here.
>
> The final model by manual backward AIC selection is `status_binary ~ thickness + ulcer`.

### Question 10

Repeat this process using the step() function.

```{r}
#| label: step-aic

M_step <- step(M4, direction = "backward", trace = 1)
```

```{r}
#| label: final-step-model

summary(M_step)
```

> `step()` arrives at exactly the same final model, `status_binary ~ thickness + ulcer`, confirming the manual process above.

### Question 11

For your final model, provide clinical interpretations of the estimated coefficients on the odds-ratio scale. Your interpretations should be written in language accessible to a clinician, not a statistician. Note any limitations that should be communicated with clinicians and/or patients.

```{r}
#| label: final-model

final_model <- glm(status_binary ~ thickness + ulcer,
                   data = melanoma,
                   family = binomial())

exp(coef(final_model))
```

```{r}
#| label: final-model-ci

exp(confint(final_model))
```

> **Warning — Statistical Associations**
>
> Recall that we are finding associations, so be careful not to interpret them as causation.
>
> **Tumour thickness:** For every extra 1 mm of tumour thickness at the time of surgery, a patient's odds of dying due to melanoma are about **14% higher** (odds ratio `1.14`, 95% CI `1.01` to `1.27`), after accounting for whether the tumour was ulcerated.
>
> **Ulceration:** Patients whose tumour was ulcerated had almost **4 times the odds** of dying from melanoma compared with patients with a non-ulcerated tumour of the same thickness (odds ratio `3.87`, 95% CI approximately `1.89` to `8.07`).
>
> **Limitations**
>
> -   These are odds ratios, not risk ratios or survival probabilities. They should not be presented to a patient as if they mean “your risk of dying is 4x higher”; they are population-level statistical associations.
>
> -   It is unclear how long patients were followed for. The outcome ignores how long each patient was followed. A more appropriate model would use time-to-event methods properly if follow-up time information were available.
>
> -   Patients who died from unrelated causes are coded as `0` alongside true survivors. This may be misleading because they may have eventually died from melanoma, but the data does not reflect this.
>
> -   This model was built and evaluated on the same set of 205 patients, with no separate validation set. Therefore, its predictive performance may not generalise as well to new patients as the in-sample statistics suggest.

## References

Gershenwald, Jeffrey E., Richard A. Scolyer, Kenneth R. Hess, et al. 2017. “Melanoma Staging: Evidence-Based Changes in the American Joint Committee on Cancer Eighth Edition Melanoma Staging Manual.” *CA: A Cancer Journal for Clinicians* 67 (6): 472–92. https://doi.org/10.3322/caac.21409.
